### 2023-02-01 tex/rensub.tex
[ writing ] more

\chapter{Usage restriction via semirings}\label{sec:semirings}

The methods described in \cref{sec:simple} for the simply typed
$\lambda$-calculus
make crucial use of \emph{weakening} --- the fact that if we have
$\Gamma \vdash A$, then we also have $\Gamma, \Delta \vdash A$.
We use this property to update environments as we take them under binders.
However, as we saw in \cref{sec:linearity}, there are interesting calculi in
which general weakening does not hold.
As such, one of the aims of this chapter will be to find a form of weakening
applicable to variables of any type, while essentially retaining linearity
(as opposed to affineness).

As explained by \citet{McBride16}, the first insight is to, instead of
removing variables from the context of certain subterms, add an annotation to
free variables saying whether or not they are to be used.
I use an annotation $\gr0$ on variables that are not to be used, and an
annotation $\gr1$ on variables that are to be used.
This convention lets us transcribe the usual $\otimes$-introduction rule
(below left) as a rule with usage annotations (below right).
In the notation on the right, I let $\Gamma = \grP\gamma$ and
$\Delta = \grQ\delta$, where $\Gamma$ is a whole context comprising a
\emph{usage context} $\grP$ and a \emph{typing context} $\gamma$.
A usage context is a list of usage annotations, so
$\grP = \gr{r_1}, \ldots, \gr{r_m}$ and a typing context is a list of types, so
$\gamma = A_1, \ldots, A_m$.
When combined, the usage context and the typing context will be of the same
length.
Explicit contexts will usually be written with usage annotations and types
interspersed, as $\gr{r_1}A_1, \ldots, \gr{r_m}A_m$.
I use $\gr r\gamma$ to abbreviate $\gr rx_1, \ldots, \gr rx_m$.

\[
  \ebrule{%
    \hypo{\Gamma \vdash A}
    \hypo{\Delta \vdash B}
    \infer2{\Gamma, \Delta \vdash A \otimes B}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \hypo{\gr1\gamma, \gr0\delta \vdash A}
    \hypo{\gr0\gamma, \gr1\delta \vdash B}
    \infer2{\gr1\gamma, \gr1\delta \vdash A \otimes B}
  }
\]

The eventual target of all these $\gr0$-annotated variables is the variable
rule, which I transcribe as follows.
The $\gr1$ shows us that we can use the variable thus annotated, while the
$\gr0$s let us discard all of the other variables in $\gamma$.

\[
  \ebrule{%
    \infer0{A \vdash A}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \infer0{\gr0\gamma, \gr1A \vdash A}
  }
\]

The use of $\gr0$ gives us the property that variables never go out of
scope in subterms; rather, we lose the ability to use certain variables, but
retain the ability to refer to them metatheoretically.
Additionally, we recover a form of weakening: if $\Gamma \vdash A$, then also
$\Gamma, \gr0\delta \vdash A$, because the resulting term indeed uses no
variables from $\delta$.
I prove the admissibility of weakening for terms will come in \cref{sec:lrsub}.

If we follow the DILL style of variable management explained in \cref{sec:dill},
there are not just the two
states \emph{to be used} ($\gr1$) and \emph{not to be used} ($\gr0$), but also
\emph{usable unrestrictedly}.
If we assign unrestricted (or \emph{intuitionistic}) variables an annotation
$\gr\omega$, we can make the following transcription of the DILL
$\otimes$-introduction rule.

\[
  \ebrule{%
    \hypo{\Theta; \Gamma \vdash A}
    \hypo{\Theta; \Delta \vdash B}
    \infer2{\Theta; \Gamma, \Delta \vdash A \otimes B}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
  }
\]

To conceptualise the criteria on the usage annotations involved in this rule,
I introduce an additive structure over usage annotations.
The rule stated above relies on the facts that $\gr1 + \gr0 = \gr1$,
$\gr0 + \gr1 = \gr1$, and $\gr\omega + \gr\omega = \gr\omega$.
Addition lifts pointwise to vectors of usage annotations (the green capital
calligraphic $\grP$, $\grQ$, and $\grR$).
A beneficial side-effect of the fact that $\gr0 + \gr0 = \gr0$ is that the
rule on the right below is actually more general, and accepts $\gr0$-annotated
variables in its
conclusion, which is essential for weakening to be admissible.

\[
  \ebrule{%
    \hypo{\gr\omega\theta, \gr1\gamma, \gr0\delta \vdash A}
    \hypo{\gr\omega\theta, \gr0\gamma, \gr1\delta \vdash B}
    \infer2{\gr\omega\theta, \gr1\gamma, \gr1\delta \vdash A \otimes B}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \hypo{\grR = \grP + \grQ}
    \hypo{\grP\gamma \vdash A}
    \hypo{\grQ\gamma \vdash B}
    \infer3{\grR\gamma \vdash A \otimes B}
  }
\]

Some other transcriptions from DILL to the usage annotation style are as
follows.
I unify the variable rules (the one for linear variables and the one for
intuitionistic variables) by introducing a coercibility ordering $\leq$ on usage
annotations.
We have $\gr\omega \leq \gr1$ because an intuitionistic variable can fill the
demand of a linear variable by dereliction.
We also have $\gr\omega \leq \gr0$, because intuitionistic variables can be
weakened away like $\gr0$-annotated variables.
This ordering information is shown in the diagram
\begin{tikzpicture}[baseline]
  \node(omega) at (0,0) {$\gr\omega$};
  \node(0) [above left of=omega] {$\gr0$};
  \node(1) [above right of=omega] {$\gr1$};

  \draw (omega) -- (0);
  \draw (omega) -- (1);
\end{tikzpicture}.
All together, this means that at the (only) variable rule, the variable being
used must have annotation less than or equal to $\gr1$, and every other variable
must have annotation less than or equal to $\gr0$.
I write this requirement as $\grR \leq \langle x \rvert$, where
$\langle x \rvert$ is the \emph{basis vector} at position $x$.

\[
  \ebrule{%
    \infer0{\Theta; A \vdash A}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \infer0{\gr\omega\theta, \gr1A, \gr0\delta \vdash A}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \hypo{\grR \leq \bra x}
    \hypo{\gamma_x = A}
    \infer2{\grR\gamma \vdash A}
  }
\]

\[
  \ebrule{%
    \infer0{\Theta, A; {\cdot} \vdash A}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \infer0{\gr\omega\theta, \gr\omega A, \gr0\delta \vdash A}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \hypo{\grR \leq \bra x}
    \hypo{\gamma_x = A}
    \infer2{\grR\gamma \vdash A}
  }
\]

The final interesting rule form to cover is found in DILL's
$\oc$-introduction rule.
DILL's $\oc$-introduction can be though of as an $\gr\omega$-ary counterpart to
$\otimes$-introduction, though with the same premise each time rather than
$\gr\omega$-many premises.
This explains why only $\gr\omega$- and
$\gr0$-annotated variables can appear in the conclusion of $\oc$-introduction,
and also justifies the choice below of multiplication (vector scaling) as the
algebraic operation controlling the $\oc$-modality.

\[
  \ebrule{%
    \hypo{\Theta; {\cdot} \vdash A}
    \infer1{\Theta; {\cdot} \vdash \oc A}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \hypo{\gr\omega\theta, \gr0\delta \vdash A}
    \infer1{\gr\omega\theta, \gr0\delta \vdash \oc A}
  }
  \quad\rightsquigarrow\quad
  \ebrule{%
    \hypo{\grR \leq \gr\omega\grP}
    \hypo{\grP\gamma \vdash A}
    \infer2{\grR\gamma \vdash \oc_{\gr\omega} A}
  }
\]

In summary, the structure we have required of the set of usage annotations is
that they have addition (for $\otimes$-introduction and similar rules),
multiplication (for $\oc$-introduction), a $1$ (for a variable being used), a
$0$ (for a variable not being used), and an ordering (allowing for subsumption
of usage restrictions).
Together, these form a \emph{partially ordered semiring} (posemiring), the laws
of which are both supported by examples and necessary for the syntax to be well
behaved.
%\todo{Refer back to preliminaries for definition of posemiring}
For concreteness, I collect together the definition of the
$\{\gr0, \gr1, \gr\omega\}$ posemiring I have been using so far.

\begin{example}\label{def:lin-semiring}
  The \emph{$\{\gr0, \gr1, \gr\omega\}$ posemiring}, also known as the
  \emph{linearity posemiring}, has the operations given as follows, with
  $0 \coloneqq \gr0$ and $1 \coloneqq \gr1$:

  \makebox[\textwidth][s]{
    \begin{tabular}{c|ccc}
      $+$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
      $\gr0$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\
      $\gr1$ & $\gr1$ & $\gr\omega$ & $\gr\omega$ \\
      $\gr\omega$ & $\gr\omega$ & $\gr\omega$ & $\gr\omega$ \\
    \end{tabular}
    \begin{tabular}{c|ccc}
      $*$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\ \hline
      $\gr0$ & $\gr0$ & $\gr0$ & $\gr0$ \\
      $\gr1$ & $\gr0$ & $\gr1$ & $\gr\omega$ \\
      $\gr\omega$ & $\gr0$ & $\gr\omega$ & $\gr\omega$ \\
    \end{tabular}
    \begin{tikzpicture}[baseline]
      \node(omega) at (0,0) {$\gr\omega$};
      \node(0) [above left of=omega] {$\gr0$};
      \node(1) [above right of=omega] {$\gr1$};

      \draw (omega) -- (0);
      \draw (omega) -- (1);
    \end{tikzpicture}
  }
\end{example}

%In \cref{sec:simple}, we saw that the logical rules of simply typed
%$\lambda$-calculus can be described in terms of three basic premise combinators:
%$\dot1$, standing for no premises; $\dottimes$, allowing for multiple premises
%in the same context; and $\Theta \vdash A$, requiring a subterm of type $A$
%having bound the extra variables from context $\Theta$.
%However, we remember from \cref{sec:linearity} that in a substructural setting,
%we do not always want to copy assumptions for use in all subterms.
%This motivates me to introduce the additional premise combinators $I^*$, $\sep$,
%and $\cdot$ in \cref{sec:lnd}, allowing for the modes of usage exhibited in
%the introduction rules for $I$, $\otimes$, and $\oc$, respectively.

The rest of this chapter proceeds as follows.
In \cref{sec:lr}, I define the usage-annotated calculus \name{}, which has
appeared in my previous work~\citep{WA21}, and can be seen as a
simply typed version of Atkey's dependently typed calculus QTT~\citep{Atkey18}.
The idea of \name{} is to augment the simply typed $\lambda$-calculus with
annotations on free variables, and give enough types to manipulate these
annotations.
Given this new calculus $\name$, the first goal is to apply the techniques of
\cref{sec:simple} to it, yielding a simultaneous substitution operation.
To do this, I use \cref{sec:lnd} to introduce notation that allows us to restate
the typing rules of \name{} to not mention contexts explicitly, as was the style
in \cref{sec:simple}.
This new notation --- the \emph{bunched connectives} --- is versatile at
defining simply typed usage-aware syntaxes, and I give further non-$\name$
examples in \cref{sec:rec}.
Finally, I justify connections to linear logic and modal logic in
\cref{sec:translation}, where I translate $\name$ terms to and from
DILL~\citep{Barber1996} and the modal calculus of \citet{judgmental}.
%The application of the techniques of \cref{sec:simple} is left to
%\cref{sec:ren-sub-lr}.

\section{A usage-annotated calculus $\name$}\label{sec:lr}
\input{lr.tex}
\section{Bunched connectives}\label{sec:lnd}
\input{lnd.tex}
\section{Additions to and variations of \name{}}\label{sec:variant}
\input{rec.tex}
\section{Translation to and from existing systems}\label{sec:translation}
\input{trans.tex}
\section{Addendum: (lack of) partiality}\label{sec:part}
\input{part.tex}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Conclusion}\label{sec:ren-sub-lr-conc}

In the preceding two chapters, I have developed a discipline for specifying
the syntax of linear and modal type systems, and furthermore developing the
syntactic metatheory of those type systems.
All of these are based on semirings, and the linear algebra arising from
considering a usage context full of semiring elements as a vector.

@@

A similar substitution lemma appears in the PhD thesis of
\citet[p.\ 138]{petricek-thesis} under the name \emph{multi-nary substitution}.
In my notation, \citeauthor{petricek-thesis}'s substitution rule looks like the
following, up to permutation of the contexts containing $\Gamma$.
Note that if $\Delta = \grQ\delta$, then $\gr r\Delta$ denotes the context
$\plr{\gr r\grQ}\delta$.
This rule is essentially an iterated version of the standard linear single
substitution principle, and is used by \citeauthor{petricek-thesis} as a
strengthened induction hypothesis required to derive single substitution.

\[
  \ebrule{%
    \hypo{\Delta_1 \vdash A_1}
    \hypo{\cdots}
    \hypo{\Delta_n \vdash A_n}
    \hypo{\Gamma, \gr{r_1}A_1, \ldots, \gr{r_n}A_n \vdash B}
    \infer4{\Gamma, \gr{r_1}\Delta_1, \ldots, \gr{r_n}\Delta_n \vdash B}
  }
\]

We can derive \citeauthor{petricek-thesis}-style multi-nary substitution as a
corollary of my simultaneous substitution, using reasoning similar to that of
\cref{thm:single-sub}.

\begin{corollary}\label{thm:petricek-sub}
  \Citeauthor{petricek-thesis}'s multi-nary substitution, as stated above, is
  admissible in $\name$.
\end{corollary}
\begin{proof}
  It is enough to provide a substitution of type
  \[
    \Gamma, \gr{r_1}\Delta_1, \ldots, \gr{r_n}\Delta_n
    \env\vdash \Gamma, \gr{r_1}A_1, \ldots, \gr{r_n}A_n.
  \]
  To do this, we use \cref{thm:construct-env} repeatedly, leaving us needing a
  substitution of type
  $\Gamma, \gr0\Delta_1, \ldots, \gr0\Delta_n \env\vdash \Gamma$ and terms of
  types
  \begin{align*}
    \gr0\gamma, \Delta_1, \gr0\delta_2, &\ldots, \gr0\delta_{n-1}, \gr0\delta_n
    \vdash A_1 \\
    &\vdots \\
    \gr0\gamma, \gr0\delta_1, \gr0\delta_2, &\ldots, \gr0\delta_{n-1}, \Delta_n
    \vdash A_n.
  \end{align*}
  The identity substitution and weakening by $\gr0$-annotated variables is
  enough to make these requirements line up with the given hypotheses.
\end{proof}

My substitution principle is stronger than \citeauthor{petricek-thesis}'s.
Where \citeauthor{petricek-thesis} requires that distinct variables be
available for each hypothesis, I allow for separation of uses via addition of
contexts.
Below is a prototypical example.

\begin{example}
  Let $\Ann \coloneqq \plr{\mathbb N, =, 0, +, 1, \times}$, the exact
  usage-counting posemiring.
  Then, we can construct a substitution $\rho : \gr2A \env\vdash \gr1A, \gr1A$,
  yielding a transformation of terms of the following form:
  \[
    \ebrule{%
      \hypo{\gr1A, \gr1A \vdash B}
      \infer1{\gr2A \vdash B}
    }.
  \]
  To construct $\rho$, we use \cref{thm:construct-env} case
  $\sep(\rightarrowtriangle)$, using the fact that $\gr2 \leq \gr1 + \gr1$.
  From there, two identity substitutions suffice.
  The action of $\rho$ on terms is to merge the two variables into one.
  Note that a renaming, rather than a substitution, would also suffice.
\end{example}

Most notably, my (single) substitution principle more naturally fits the
requirement we would have for the reduct of the $\beta$-rule for functions in
$\name$, whereas \citeauthor{petricek-thesis}'s substitution principle would
need some additional transformation for it to fit properly.
This comes from the fact that the $\name$ function application rule introduces
an algebraic ($+$) separation between its premises, whereas
\citeauthor{petricek-thesis}'s substitution principle separates premises only
via concatenation.

>>>>

@@

\section{What are linear renaming and substitution?}\label{sec:lrkits}
\input{lrkits.tex}
\section{Properties of linear environments}\label{sec:lenv}
\input{lenv.tex}
\section{Substitution is admissible in \name{}}\label{sec:lrsub}
\input{lrsub.tex}
\section{Comparison with Petricek's substitution lemma}\label{sec:petricek}
\input{petricek.tex}
\section{Conclusion}\label{sec:ren-sub-lr-conc}

In this and the preceding chapter, I have developed a discipline for specifying
the syntax of linear and modal type systems, and furthermore developing the
syntactic metatheory of those type systems.
All of these are based on semirings, and the linear algebra arising from
considering a usage context full of semiring elements as a vector.

@@

\begin{itemize}
  \item Everything presented in this chapter is perfect.
  \item But equivalence of substitutions is hard to define/work with.
  \item The process of this development gives a template for defining
    simultaneous substitution for other substructural calculi.
\end{itemize}

I have made few unforced design decisions in this chapter, so there are few
aspects I can see future work improving upon.
However, future work may want to extend the work of this chapter, in which case
there are some unanswered questions.
Principle among these, in my mind, is dealing with equivalence/equality of
environments.
Intuitionistic environments $\rho, \sigma : \Gamma \env\V \Delta$ are equal
if and only if, for each type $A$ and each variable $x : \Delta \ni A$, we have
$\rho\,x = \sigma\,x$.
This follows from what we expect of equality of functions (function
extensionality).
Usage-aware environments, on the other hand, are $\Sigma$-types --- a way of
dividing up the usage annotations of $\Gamma$, and then a function producing
$\V$-values whose usage annotations come from that division.
Equality of $\Sigma$-types is tricky --- we need to equate the first components,
then rewrite the types of the second components by this equation before equating
them.
In practice, the equations rewriting other other equations build up so much that
I have given up on a first effort to give a treatment of the equational theory
of substitutions.
Note that recursively defined environments (\cref{def:lr-rec-env}) are also
$\Sigma$-types in cases where $\Delta$ is a concatenation of contexts, so that
definition does not clearly help.

### 2023-02-01 tex/semantics.tex
[ writing ] more

Given a $\V$-environment $\Gamma \Rightarrow \Delta$, the function
\AgdaFunction{semantics} we define in this section assigns a
$\C$-value in context $\Gamma$ to every term in context $\Delta$,
where $\C$ is an \AgdaFunction{OpenFam} being the carrier of the
semantic interpretation of terms ($\V$ being the semantic
interpretation of variables). Before we can define
\AgdaFunction{semantics}, we need to treat recursion through rules'
premises (\cref{sec:functorial}) and extension of environments when
going under variable binders (\cref{sec:kripke}).

@@

\subsection{A layer of syntax is functorial}\label{sec:functorial}

A basic property of the universe of syntaxes we described in \cref{sec:syntax}
is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
Its type says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{F}.

@@

The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
works through the structure \AgdaBound{ps} of premises of the rule applied,
acting on each subterm it finds.
Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
and \AgdaBound{Q$'$}.
In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
The cases for \AgdaInductiveConstructor{`$\cdot$} and
\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
of linearity.
In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
systems, do not involve any changes in the usage contexts.

@@

\subsection{The Kripke function space}\label{sec:kripke}

At this point we introduce a minor generalisation to
\AgdaFunction{OpenFam} and \AgdaFunction{ExtOpenFam}:
\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---ExtOpenFam}.  We obtain the
definition of \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} by
replacing the textual occurrence of \AgdaBound{Ty} by the parameter
\AgdaBound{I}.

The definition \AgdaFunction{Kripke}\,$\V$\,$\C$\,$\Delta$ is a kind
of function space that describes a $\C$ value parametrised by
$\Delta$-many additional $\V$s (all correctly typed and usage
annotated). It is used to describe how to go under binders in a
Higher-Order Abstract Syntax style---to go under a binder we must
provide semantic interpretations for all the additional variables:

@@

If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
\ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.  That is
to say, the Kripke function is expecting a value for each newly bound
variable, at the multiplicity of its annotation, together with the
resources supporting each of those values. We use the ``magic wand''
function space here to enforce the invariant that the freshly bound
variables have usage annotations that are added to the existing
variables, not shared with them. The use of the
\AgdaFunction{$\Box^r$} modality ensures that we can still use it in
the presence of additional variables introduced by weakening.

>>>>

Having fixed a universe of syntaxes, in which we can build terms, the next thing
to do is to write recursive functions on terms.
With terms being given by an inductive \AgdaSymbol{data} type definition, they
already come with a recursion and an induction principle.
However, these principles do not handle variable-binding, which we have seen
with the fact that we had to write the \AgdaFunction{bind} helper
function for renaming and substitution in \cref{sec:lrsub}.

In this chapter, the central construct is a function \AgdaFunction{semantics}
which, for a $\V$-environment $\rho : \Gamma \env\V \Delta$,
maps a term $M : \Delta \vdash A$ to some semantic value in the type
$\C\,\Gamma\,A$.
This is a direct adaptation of the \AgdaFunction{semantics} function of
\cref{sec:gen-sem}, which has the same kind of action on intuitionistic terms,
given similar operations on $\V$ and $\C$ as what we had earlier.
The \AgdaFunction{semantics} function recurses on the term $M$, updating $\rho$
whenever new variables are bound.
In our usage-aware case, $\rho$ is also updated whenever we come across linear
combinations induced by premise combinators $I^{\sep}$, $\sep$, and
$\gr r \cdot {}$.

This chapter is structured as follows.
I start by giving a quick introduction to linear relations --- a generalisation
of linear maps --- with reference to their use in mechanised algebraic
reasoning, in \cref{sec:lin-rel}.
Using linear relations, I give a functorial \emph{map} operation to a single
layer of syntax in \cref{sec:functorial}.
I then adapt the \AgdaFunction{Kripke} function space to the usage-aware
setting in \cref{sec:kripke}.
Then I apply the \AgdaFunction{Kripke} function space, along with much of the
machinery I have introduced in previous chapters, to give the
\AgdaFunction{semantics} function in \cref{sec:traversal}.
Finally, I give some example uses of \AgdaFunction{semantics} in
\cref{sec:example-semantics}.
\todo{What are the examples?}

@@

\section{Linear relations in Agda}\label{sec:lin-rel}

In \cref{sec:lrkits}, I defined \emph{usage-annotated environments}
(\cref{def:lr-env}).
One component of a usage-annotated environment is a linear map $\gr\Psi$ which,
when applied to the target usage vector, gives a vector compatible with the
source usage vector.

When it comes to mechanisation, I prefer to replace an assertion
``$\grP \leq \grQ\gr\Psi$'', involving a linear map $\gr\Psi$, by an assertion
``$\grP\gr\Psi\grQ$'', where $\gr\Psi$ is now a linear \emph{relation} said to
relate $\grP$ and $\grQ$.
I define linear relations as follows, where the reader may wish to check that
a linear map gives rise to a linear relation via the expression
$\grP \leq \grQ\gr\Psi$.

\begin{definition}\label{def:linear-relation}
  Given a posemiring $\Ann$ and modules $\mathscr M$ and $\mathscr N$ over
  $\Ann$, a \emph{linear relation} between $\mathscr M$ and $\mathscr N$ is
  a relation $\gr\Psi$ between the underlying sets of $\mathscr M$ and
  $\mathscr N$ such that the following properties hold of all
  $\grP, \grPprime, \grPl, \grPr \in \mathscr M$ and all
  $\grQ, \grQprime, \grQl, \grQr \in \mathscr N$.
  \begin{align*}
    \grPprime \leq \grP \land \grP\gr\Psi\grQ \land \grQ \leq \grQprime
    &\implies \grPprime\gr\Psi\grQprime
    \\
    \plr{\exists\grQ.~\grP\gr\Psi\grQ \land \grQ \leq \gr0}
    &\implies \grP \leq \gr0
    \\
    \plr{\exists\grQ.~\grP\gr\Psi\grQ \land \grQ \leq \grQl + \grQr}
    &\implies \plr{\exists\grPl,\grPr.~\grP \leq \grPl + \grPr
      \land \grPl\gr\Psi\grQl \land \grPr\gr\Psi\grQr}
    \\
    \plr{\exists\grQ.~\grP\gr\Psi\grQ \land \grQ \leq \gr r\grQprime}
    &\implies \plr{\exists\grPprime.~\grP \leq \gr r\grPprime
      \land \grPprime\gr\Psi\grQprime}
  \end{align*}
  I write $\mathscr M \rel \mathscr N$ as the type of linear relations between
  $\mathscr M$ and $\mathscr N$.
\end{definition}

Relations have several advantages over functions when doing mechanised algebra
in type theory.
For one, what are compound expressions in functional style --- for example
$x \leq f(y) + g(z)$ --- become collections of simple relationships in
relational style --- for example $\exists v,w.~vfy \land wgz \land
\mathrm{Add}\,x\,v\,w$.
The advantage of this is that we have immediate access to all of the expressions
and subexpressions, and the proofs of the relationships between them.
This means that there is no need for congruence or monotonicity lemmas, and
correspondingly no need to explicitly describe the syntactic context in which
some algebraic manipulation is being applied and we rely less on the unifier.
Another advantage is that one can design relations so that pattern-matching
suggestively decomposes complex relationships.
For example, given $F : \mathscr M \rel \mathscr M'$ and
$G : \mathscr N \rel \mathscr N'$, we can define a relation
$F \oplus G : \mathscr M \oplus \mathscr N \rel \mathscr M' \oplus \mathscr N'$
pointwise, so that a proof of $(x, x')(F \oplus G)(y, y')$ is a proof of $xFy$
together with a proof of $x'Gy'$.
Pattern-matching on such a proof immediately gives us these constituent parts,
whereas proofs of the corresponding statement involving functions would require
using a lemma to get the parts.
There is a dual advantage when producing one of these proofs, where we can
introduce the canonical constructor (for pairs, in this example) rather than
having to find the appropriate lemma.

Relations also have several disadvantages, though I have found that for my use
case, these are outweighed by the advantages.
For example, automated algebraic solvers are better developed for function-based
algebraic expressions, and sometimes the fact that functions satisfy unitality
and associativity up to decidable judgemental equality means that some proofs
can be avoided.
The handling of compound expressions can also be a disadvantage in that it
necessitates lots of new variable names and obscures goal and context displays.
Finally, in predicative systems such as Agda, relations typically live in a
larger universe than the corresponding functions.
In practice, this means quantifying over an extra level variable for each
relation involved in general lemmas.

There are more relations than there are functions, so statements involving
relations are more general than the corresponding statements involving
functions.
However, one part of the development requires functions rather than relations,
so I impose functionality on relations after the fact.
The appropriate notion of functional relation I use is slightly different to the
standard one, in that I take account of the order on the codomain, and thus ask
for the \emph{largest} solution rather than the \emph{unique} solution.

\begin{definition}\label{def:functional-linear-relation}
  A linear relation $\gr\Psi$ between $\mathscr M$ and $\mathscr N$ is
  \emph{(right-to-left) functional} if, for every $\grQ \in \mathscr N$, there
  exists universally a $\grP \in \mathscr M$ such that $\grP\gr\Psi\grQ$.
  Universality means that, for all $\grPprime$ such that $\grPprime\gr\Psi\grQ$,
  we have $\grPprime \leq \grP$ (i.e.\ $\grP$ is the largest solution).
\end{definition}

In Agda code, $\gr\Psi$ becomes \AgdaBound{$\Psi$} and the fact that $\gr\Psi$
relates $\grP$ and $\grQ$ (in this section written $\grP\gr\Psi\grQ$) is
rendered as
\PsiDot{rel}\AgdaSpace{}\AgdaBound{P}\AgdaSpace{}\AgdaBound{Q}.
That $\gr\Psi$ respects the orders on its arguments is given by
\PsiDot{rel-$\leq_m$}, and the various linearity properties are given by
\PsiDot{rel-0$_m$}, \PsiDot{rel-+$_m$}, and \PsiDot{rel-*$_m$}.

\section{A layer of syntax is functorial}\label{sec:functorial}

A basic property of the universe of syntaxes
is that every syntax supports a functorial action on subterms, realised by a
function \AgdaFunction{map-s}.
Its type says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{$\Psi$} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{$\Psi$}.

@@

The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
works through the structure \AgdaBound{ps} of premises of the rule applied,
acting on each subterm it finds.
Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
and \AgdaBound{Q$'$}.
In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
Linearity of \AgdaBound{$\Psi$} yields $\grQ_M$ and $\grQ_N$ such that
$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaFunction{map-p} recursively at
$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
The cases for \AgdaInductiveConstructor{`$\cdot$} and
\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
of linearity.
In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
systems, do not involve any changes in the usage contexts.

The linearity of relation \AgdaBound{$\Psi$} is given by fields
\AgdaField{rel-0$_m$}, \AgdaField{rel-+$_m$}, and \AgdaField{rel-*$_m$} (with
the subscript-m being a mnemonic for \emph{module}, as opposed to scalar).
\todo{Refer to preliminaries}

@@

I have also extended \AgdaFunction{map-p} to handle the various
$\Box$-modalities described in \cref{sec:dup-lnl}.
The Agda code for this extension is not particularly readable, so I do not
include it in this document.
However, this extension is notable as the only part of the framework requiring
that the linear relation \AgdaBound{$\Psi$} be functional (i.e., total and
deterministic).

\section{The Kripke function space}\label{sec:kripke}

At this point I introduce a minor generalisation to
\AgdaFunction{OpenFam} and \AgdaFunction{ExtOpenFam} (as defined in
\cref{sec:terms}):
\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---ExtOpenFam}.  We obtain the
definition of \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} by
replacing the textual occurrence of \AgdaBound{Ty} by the parameter
\AgdaBound{I}, though there is still reference to the ambient \AgdaBound{Ty}
via \AgdaRecord{Ctx}.
The main value I am interested in \AgdaBound{I} taking, other than
\AgdaBound{Ty}, is \AgdaRecord{Ctx} --- for example, the type family of
$\V$-environments, for a given $\V$, is a
\AgdaRecord{Ctx}\AgdaSpace{}\AgdaFunction{---OpenFam}%
\AgdaSpace{}\AgdaSymbol{\_}.
However, I only use this generalisation much later.
\todo{When?}

\ExecuteMetaData[\Syntaxtex]{dashOpenFam}

The definition
\AgdaFunction{Kripke}\AgdaSpace{}\AgdaBound{$\V$}\AgdaSpace{}\AgdaBound{$\C$}%
\AgdaSpace{}\AgdaBound{$\Delta$} is a kind
of function space that describes a \AgdaBound{$\C$}-value parametrised by
\AgdaBound{$\Delta$}-many additional \AgdaBound{$\V$}-values (all correctly
typed and usage-annotated).
It is used to describe how to go under binders in a
Higher-Order Abstract Syntax style: To go under a binder we must
provide semantic interpretations for all the additional variables.

@@

If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
\ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.  That is
to say, the Kripke function is expecting a value for each newly bound
variable, at the multiplicity of its annotation, together with the
resources supporting each of those values. We use the ``magic wand''
function space here to enforce the invariant that the freshly bound
variables have usage annotations that are added to the existing
variables, not shared with them. The use of the
\AgdaFunction{$\Box^r$} modality ensures that we can still use it in
the presence of additional variables introduced by weakening.
\todo{Is wand defined?}
